---
title: "Module 3: Pharmacogenomics for biomarker discovery – Basic analysis"
author: 
  - name: Jermiah J. Joseph
    email: jermiah.joseph@uhn.ca
    affiliation:
        - &pm Princess Margaret Cancer Centre
  - name: Nikta Feizi
    email: 
    affiliation:
        - &pm Princess Margaret Cancer Centre
  - name: Julia Nguyen
    email: julia.nguyen@uhn.ca
    affiliation:
        - &pm Princess Margaret Cancer Centre
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('CBWWorkshop2024')`"
vignette: >
  %\VignetteIndexEntry{Module 3: Intro to pharmacogenomics in the context of cancer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=TRUE, echo=FALSE}
PGx <- BiocStyle::Biocpkg("PharmacoGx")
CGx <- BiocStyle::Biocpkg("CoreGx")
TGx <- BiocStyle::Biocpkg("ToxicoGx")
RGx <- BiocStyle::Biocpkg("RadioGx")
Xv <- BiocStyle::Biocpkg("Xeva")
DT <- BiocStyle::CRANpkg("data.table")
SE <- BiocStyle::Biocpkg("SummarizedExperiment")
MAE <- BiocStyle::Biocpkg("MultiAssayExperiment")
```

## Overview 
# Lab 3 Overview {-}

## Instructor(s) name(s) and contact information {-}

* Jermiah J. Joseph <jermiah.joseph@uhn.ca>
* Nikta Feizi <>
* Julia Nguyen <julia.nguyen@uhn.ca>

## PharmacoGx Package Overview

**PharmacoGx:** An R package for analyzing pharmacogenomic datasets.

**Key Functions:**

- **Install and Load PharmacoGx**
  ```r
  install.packages("BiocManager")
  BiocManager::install("PharmacoGx")
  library(PharmacoGx)
  ```

- **Download a PharmacoSet**
  ```r
  availablePSets()
  GDSC <- downloadPSet("GDSC_2020(v2-8.2)")
  # downloadPSet(): Download pharmacogenomic datasets.
  ```

- **Extract Drug Response Data**
  ```r
  drug_response <- summarizeSensitivityProfiles(GDSC, sensitivity.measure='aac_recomputed')
  # summarizeSensitivityProfiles(): Summarize drug response data.
  ```

- **Extract Gene Expression Data**
  ```r
  gene_expr <- summarizeMolecularProfiles(GDSC, mDataType='rna') 
  gene_expr_mtx <- assay(gene_expr)
  ```

**Data Retrieval:**

- **availablePSets()**: Lists available pharmacogenomic datasets.

- **Downloading Multiple PharmacoSets:**
  ```r
  PSet.list <- lapply(c("GDSC", "CCLE", "CTRP"), downloadPSet)
  names(PSet.list) <- c("GDSC", "CCLE", "CTRP")
  ```

- **Accessing Molecular Profiles:**
  ```r
  molecular.data <- lapply(PSet.list, function(pset) {
    summarizeMolecularProfiles(pset, mDataType = 'rna')
  })
  ```

## Introduction

In this workshop, we will explore data integration and comparative analysis
using the `r PGx` package, focusing on the **GDSC** and **CCLE** datasets.
Specifically, we will cover drug sensitivity comparisons using AUC and IC50
measures and investigate correlations between gene expression profiles. This R
Markdown file is designed to give participants hands-on experience with data
exploration and interpretation.

### Learning Objectives

1. Learn how to load and handle data from the **GDSC** and **CCLE** datasets.
2. Understand how to summarize drug sensitivity profiles and molecular data.
3. Perform correlations to assess concordance between cell line datasets.
4. Visualize and interpret the results of statistical tests.

```{r load_packages, eval=TRUE, echo=FALSE, include=FALSE}
suppressPackageStartupMessages({
  library(CBWWorkshop2024)
  library(PharmacoGx)
  library(data.table)
  library(reshape2)
  library(ggplot2)
  library(ggpubr)
  library(Biobase) # Core Bioconductor class for gene expression data
  library(SummarizedExperiment) # Core Bioconductor class for summarized data
  library(S4Vectors) # Core Bioconductor class for handling S4 objects
  library(pander) # Package for nicely formatted tables
})
```


## Exploring GDSC and CCLE Datasets

### Load GDSC and CCLE Datasets

We will use the **GDSCsmall** and **CCLEsmall** sample datasets provided by
`r PGx` to compare the two data sources.

```{r load_data}
data("GDSCsmall")
data("CCLEsmall")
GDSCsmall
```

### Find Common Genes Between GDSC and CCLE

The first step in comparing datasets is to find shared features between them.
Here, we identify the genes that are present in both datasets.

```{r find_common_genes}
commonGenes <- intersect(fNames(GDSCsmall, "rna"), fNames(CCLEsmall, "rna"))
length(commonGenes)
```

There are 41 genes that are common between the RNA profiles of the GDSC and CCLE datasets.

### Identify Common Cell Lines and Drugs

Next, we identify the common cell lines and drugs between **GDSC** and **CCLE**.

```{r intersect_pset}
common <- intersectPSet(
  list("CCLE" = CCLEsmall, "GDSC" = GDSCsmall),
  intersectOn = c("cell.lines", "drugs"),
  strictIntersect = TRUE
)
```

There are 9 common samples between GDSC and CCLE. The `intersectPSet` function returns a 
list of the two PSets subsetted to include only the common cell lines.

```{r intersect_pset_results}
common
```

Notice that most of the molecular profiles are subsetted to include the 9 common samples. 
Some odd cases include RNA from GDSC which has a technical replicate (hence 10 samples) and
CNV from CCLE which is missing CNV profiles for two fo the nine cell lines.


### Summarize Drug Sensitivity Profiles

We summarize drug sensitivity profiles (AUC and IC50) for each dataset. The
`summary.stat` parameter can be set to different statistical metrics such as
`mean`, `median`, etc. Here, we use the `median` for summarizing the sensitivity.

```{r summarize_profiles}
# Summary statistics for AUC
GDSC.auc <- summarizeSensitivityProfiles(
  common$GDSC,
  sensitivity.measure = "auc_published",
  summary.stat = "median",
  verbose = FALSE
)
CCLE.auc <- summarizeSensitivityProfiles(
  common$CCLE,
  sensitivity.measure = "auc_published",
  summary.stat = "median",
  verbose = FALSE
)

# Summary statistics for IC50
GDSC.ic50 <- summarizeSensitivityProfiles(
  common$GDSC,
  sensitivity.measure = "ic50_published",
  summary.stat = "median",
  verbose = FALSE
)
CCLE.ic50 <- summarizeSensitivityProfiles(
  common$CCLE,
  sensitivity.measure = "ic50_published",
  summary.stat = "median",
  verbose = FALSE
)
CCLE.auc |> head()
```

We can visualize the distribution of drug sensitivity profiles to compare between the two psets. 

First we quickly reformat the data into a long format to faciliate plotting. We use
the `melt()` function from the `reshape2` package.

```{r format_sensitivity}
CCLE_toPlot <- melt(CCLE.auc)
colnames(CCLE_toPlot) <- c("Drug", "Gene", "AUC")

CCLE_toPlot |> head()
```

Next, we can create a quick box plot to visualize the AUC distributions per drug.

```{r plot_sensitivity}
ggplot(CCLE_toPlot, aes(x = Drug, y = AUC)) + geom_boxplot(fill = "grey") +
  theme_classic() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) #rotate x-axis labels
```

Let's do this again but with the GDSC PSet so we can compare the two
```{r plot_sensitivity_compare}
# format the CCLE sensitivity data
GDSC_toPlot <- melt(GDSC.auc)
colnames(GDSC_toPlot) <- c("Drug", "Gene", "AUC")

# merge the two pset dataframes
GDSC_toPlot$PSet <- "GDSC"
CCLE_toPlot$PSet <- "CCLE"
merge_toPlot <- rbind(GDSC_toPlot, CCLE_toPlot)

# plot to compare AUC distribution between PSets
ggplot(merge_toPlot, aes(x = Drug, y = AUC)) + geom_boxplot(aes(fill = PSet)) +
  scale_fill_manual(values = c("#624763", "#B1D3A3")) + theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) #rotate x-axis labels
```

Preclinical drug response data is fairly noisy, hence the variation we see between 
the two datasets. However, we do see some commonalities in distribution.

### Summarize Gene Expression Profiles

We summarize gene expression data for the genes that are common between **GDSC**
and **CCLE**.

```{r summarize_expression}
GDSCexpression <- summarizeMolecularProfiles(
  common$GDSC, cellNames(common$GDSC),
  mDataType = "rna", features = commonGenes, verbose = FALSE
)
CCLEexpression <- summarizeMolecularProfiles(
  common$CCLE, cellNames(common$CCLE),
  mDataType = "rna", features = commonGenes, verbose = FALSE
)
GDSCexpression |> head()
```

Here, you could do PCA on the profiles to quickly check the data. We will skip this 
step in the interest of time.


## Correlation Analysis Between GDSC and CCLE

We perform correlation analysis to examine the relationship between gene
expression, AUC, and IC50 measures across the **GDSC** and **CCLE** datasets
using **Spearman's correlation**.

```{r correlation_analysis_gene_expression}
# get common cell line names
cc <- cellNames(common[[1]])

# correlation of gene expression across common cell lines
ge.cor <- sapply(cc, function(x, d1, d2) {
  stats::cor(
    d1[, x], d2[, x],
    method = "spearman", use = "pairwise.complete.obs"
  )
}, d1 = assay(GDSCexpression, 1), d2 = assay(CCLEexpression, 1))

ge.cor
```

We can see that the RNA-Seq expression values are highly correlated across the 
common cell lines.

Let's take a look at the correlation of drug response, starting with IC50 values.

```{r correlation_analysis_IC50}
# quick look at the CCLE IC50 data
CCLE.ic50 |> head()

# correlation of IC50 values across common drugs and cell lines
ic50.cor <- sapply(cc, function(x, d1, d2) {
  stats::cor(
    d1[, x], d2[, x],
    method = "spearman", use = "pairwise.complete.obs"
  )
}, d1 = GDSC.ic50, d2 = CCLE.ic50)

ic50.cor
```

Notice that we have some `NA` values in the original CCLE IC50 dataset and in the
correlation results. 

The `pairwise.complete.obs` argument used in the `cor()` function ensures that only 
complete observations (i.e. numeric values that are not `NA`) are included in the 
pairwise correlations. 

Since `23132-87` and `647-V` are all `NA` values, no spearman correlation was
computed.

We can quickly compute the correlation for AUC as well.

```{r correlation_analysis_AUC}
# correlation of AUC values across common drugs and cell lines
auc.cor <- sapply(cc, function(x, d1, d2) {
  stats::cor(
    d1[, x], d2[, x],
    method = "spearman", use = "pairwise.complete.obs"
  )
}, d1 = GDSC.auc, d2 = CCLE.auc)

auc.cor
```

## Statistical Comparison

We compare the correlations using **Wilcoxon signed-rank tests** to see if there
are significant differences between the gene expression correlations and the
drug sensitivity correlations.

```{r wilcox_tests}
w1 <- stats::wilcox.test(
  x = ge.cor, y = auc.cor, conf.int = TRUE, exact = FALSE
)
w2 <- stats::wilcox.test(
  x = ge.cor, y = ic50.cor, conf.int = TRUE, exact = FALSE
)

# Display p-values
ss <- sprintf("GE vs. AUC = %.1E\nGE vs. IC50 = %.1E", w1$p.value, w2$p.value)
```

### Boxplot Visualization

The results are visualized using boxplots to compare the correlations across
gene expression, AUC, and IC50.

```{r boxplot_results}
boxplot(list("GE" = ge.cor, "AUC" = auc.cor, "IC50" = ic50.cor),
  main = "Concordance between cell lines",
  ylab = expression(R[s]),
  sub = ss,
  ylim = c(-1, 1),
  col = "lightgrey",
  pch = 20,
  border = "black"
)
```

## Drug Sensitivity Signature

We calculate drug sensitivity signatures using gene expression and mutation data
for selected drugs to understand the relationship between molecular features and
drug response. Note that this gene-drug signature is based on univariable
analysis.

```{r drug_sensitivity_signature}
features <- fNames(CCLEsmall, "rna")[which(featureInfo(CCLEsmall, "rna")$Symbol == "NQO1")]

# Gene expression signature
sig.rna <- drugSensitivitySig(
  object = CCLEsmall,
  mDataType = "rna",
  drugs = c("17-AAG"),
  features = features,
  sensitivity.measure = "auc_published",
  molecular.summary.stat = "median",
  sensitivity.summary.stat = "median",
  verbose = FALSE
)

# Mutation signature
sig.mut <- drugSensitivitySig(
  object = CCLEsmall,
  mDataType = "mutation",
  drugs = c("PD-0325901"),
  features = "BRAF",
  sensitivity.measure = "auc_published",
  molecular.summary.stat = "and",
  sensitivity.summary.stat = "median",
  verbose = FALSE
)

# Combine and display results
sig <- rbind(sig.rna, sig.mut)
rownames(sig) <- c("17-AAG + NQO1", "PD-0325901 + BRAF")
colnames(sig) <- dimnames(sig.mut)[[3]]
pander::pandoc.table(
  t(sig),
  style = "rmarkdown", caption = "P Value of Gene-Drug Association"
)
```
