---
title: "Module 4: Pharmacogenomics for biomarker discovery - Advanced analysis"
author: 
  - name: Julia Nguyen
    email: julia.nguyen@uhn.ca
    affiliation:
        - &pm Princess Margaret Cancer Centre
  - name: Nikta Feizi
    email: nikta.feizi@uhn.ca
    affiliation:
        - &pm Princess Margaret Cancer Centre
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('CBWWorkshop2024')`"
vignette: >
  %\VignetteIndexEntry{Module 4: Pharmacogenomics for biomarker discovery - Advanced analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


# Introduction and Objectives

This workshop provides an overview of meta-analysis techniques in
pharmacogenomics. By the end of this lab, participants will be able to:

- Execute a meta-analysis using multiple PharmacoSets to evaluate a biomarker of interest across different cancer types.
- Generate and interpret forest plots to visualize effect sizes.
- Perform subgroup analyses using a small subset of the CCLE dataset.
- Understand the preprocessing steps required for pharmacogenomic data.
- Assess the assumptions and model diagnostics for linear regression models.

# Load Required Libraries

```{r load_libraries, echo=TRUE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(PharmacoGx)
  library(meta)
  library(ggplot2)
  library(dplyr)
})
```

# Data Acquisition and Preparation

We will use smaller versions of the PharmacoSets for the workshop to minimize
processing time. Before diving into the analysis, let's normalize the data and
address any missing values.

```{r load-datasets}
# Load smaller datasets for demonstration purposes
data("GDSCsmall")
data("CCLEsmall")
```

# Exploratory Data Analysis (EDA)

Before performing the analysis, let's explore the data to understand its
characteristics.

```{r exploratory-data-analysis, message=FALSE, warning=FALSE}
# Summary statistics
gene <- "ENSG00000003987"

gdsc_rna_data <- summarizeMolecularProfiles(
  GDSCsmall,
  mDataType = "rna", features = gene
) |> assay()

ccle_rna_data <- summarizeMolecularProfiles(
  CCLEsmall,
  mDataType = "rna", features = gene
) |> assay()

# Distribution plots
ggplot(data = t(ccle_rna_data), aes_string(x = gene)) +
  geom_histogram(
    bins = 30, fill = "skyblue", color = "black"
  ) +
  labs(
    title = stringr::str_c("Distribution of ", gene, " Expression in CCLEsmall"),
    x = "Expression", y = "Frequency"
  )
```

# Extract Biomarker Data

Let's extract gene expression and drug response data for a biomarker of
interest, e.g., `GeneX`.

```{r extract-biomarker-data, message=FALSE, warning=FALSE}
# Extract data for biomarker
ccle_drug_data_all <- sensitivityProfiles(CCLEsmall)
ccle_drug_data_auc <- summarizeSensitivityProfiles(
  CCLEsmall,
  sensitivity.measure = "auc_published",
  summary.stat = "median",
  verbose = FALSE
)



gdsc_drug_data_all <- sensitivityProfiles(GDSCsmall)
gdsc_drug_data_auc <- summarizeSensitivityProfiles(
  GDSCsmall,
  sensitivity.measure = "auc_published",
  summary.stat = "median",
  verbose = FALSE
)
```

# Effect Size Calculation

Calculate the effect size for each dataset using linear models and check the
assumptions of the model.

```{r effect-size-calculation}
# Function to compute effect size and check model diagnostics
# Default drug and gene parameters are set for example usage
computeEffectSize <- function(study, exp_data, drug_response, drug = "Erlotinib", gene = "ENSG00000003987") {
  drug_response <- drug_response |>
    t() |>
    as.data.frame() |>
    select(all_of(drug))
  drug_response <- na.omit(drug_response) # this contains auc values

  model_data <- merge(drug_response, exp_data, by = "row.names")
  colnames(model_data) <- c("cells", "auc", "GeneExpr")

  model <- lm(auc ~ GeneExpr, data = model_data)
  coef <- summary(model)$coefficients["GeneExpr", ]
  effect <- data.frame(
    EffectSize = coef["Estimate"],
    SE = coef["Std. Error"],
    Study = study
  )
  rownames(effect) <- NULL
  return(effect)
}

# Compute effect sizes
ccle_effect <- computeEffectSize(
  study = "ccle",
  exp_data = t(ccle_rna_data),
  drug_response = ccle_drug_data_auc
)
gdsc_effect <- computeEffectSize(
  study = "gdsc",
  exp_data = t(gdsc_rna_data),
  drug_response = gdsc_drug_data_auc
)
```

# Combine Effect Sizes and Perform Meta-Analysis

We will combine the effect sizes from `GDSCsmall` and `CCLEsmall` to perform a
meta-analysis.

```{r meta-analysis}
# Combine effect sizes
effects <- rbind(ccle_effect, gdsc_effect)

# Perform meta-analysis
meta.result <- metagen(
  TE = effects$EffectSize,
  seTE = effects$SE,
  data = effects,
  sm = "SMD"
)
```

# Visualization with Forest Plots

Generate and interpret a forest plot to visualize the effect sizes.

```{r forest-plot}
# Generate forest plot
forest(meta.result)

# Enhanced forest plot with ggplot2
forest_data <- data.frame(
  Study = effects$Study,
  EffectSize = effects$EffectSize,
  LowerCI = effects$EffectSize - 1.96 * effects$SE,
  UpperCI = effects$EffectSize + 1.96 * effects$SE
)

ggplot(forest_data, aes(x = EffectSize, y = Study)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = LowerCI, xmax = UpperCI), height = 0.2) +
  geom_vline(
    xintercept = meta.result$TE.random, linetype = "dashed", color = "red"
  ) +
  labs(
    title = "Meta-Analysis of GeneX Expression and Drug Response",
    x = "Effect Size",
    y = "Study"
  ) +
  theme_minimal()
```

# Subgroup Analysis

We will now conduct a subgroup analysis using a subset of the `CCLEsmall` dataset.

```{r subgroup-analysis}
# TODO: To debug:

# Load dataset
# data("CCLEsmall")
#
# # Extract drugs and cells
# CCLEdrugs <- treatmentNames(CCLEsmall)
# CCLEcells <- sampleNames(CCLEsmall)
#
# # Subset to a specific drug and cell line
# pSet <- subsetTo(CCLEsmall, drugs = CCLEdrugs[1], cells = CCLEcells[1])
#
# # Display subset
# dim(pSet)
#
# # Subgroup analysis for tissue-specific biomarker
# gene.expr <- summarizeMolecularProfiles(pSet, mDataType = 'rna', features = gene)
# ccle_effect <- computeEffectSize(study = "ccle", exp_data = t(gene.expr) , drug_response = drug.response)
```

# Interactive Exercise

**Task**: Compare the fixed-effects and random-effects models using the effect
sizes computed above. Perform a subgroup analysis to evaluate the effect of
`GeneX` on drug response specifically for the chosen drug and cell line.

# Additional Considerations

- **Heterogeneity**: Examine the variability across different studies. Use the I² and Q-statistics to assess heterogeneity.
  
  ```{r heterogeneity}
cat("Q Statistic:", meta.result$Q, "\n")
cat("I-squared Statistic:", meta.result$I2, "%\n")
  ```

- **Limitations**: Discuss the limitations of meta-analysis, such as the variability between studies and the challenges of combining datasets from different sources.

