---
title: "Module 2: Hands-on with pharmacogenomics data"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Module 2: Hands-on with pharmacogenomics data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r eval=TRUE, echo=FALSE}
PGx <- BiocStyle::Biocpkg("PharmacoGx")
CGx <- BiocStyle::Biocpkg("CoreGx")
TGx <- BiocStyle::Biocpkg("ToxicoGx")
RGx <- BiocStyle::Biocpkg("RadioGx")
Xv <- BiocStyle::Biocpkg("Xeva")
DT <- BiocStyle::CRANpkg("data.table")
SE <- BiocStyle::Biocpkg("SummarizedExperiment")
MAE <- BiocStyle::Biocpkg("MultiAssayExperiment")
```


# Lab 2 Overview

## Instructor(s) name(s) and contact information {-}

* Jermiah J. Joseph <jermiah.joseph@uhn.ca>
* ALMAS_ DETAILS

## Lab Description {-}

<!-- Lab Practical – Hands-on with pharmacogenomics data  (Jermiah Joseph, Almas Khan)

Loading a PharmacoSet in R using PharmacoGx or ORCESTRA, understanding the data structure, accessing features, and preparing metadata
Plot batch effects using PCA and run correction method (SVA)
Filter out outliers and missing values -->

### Learning goals {-}

* Understand the data structure of a PharmacoSet
* Learn how to access features and metadata from a PharmacoSet
* Learn how to plot batch effects using PCA and run correction method (SVA)
* Learn how to filter out outliers and missing values

### Learning objectives {-}

* Describe the use cases for `r PGx` in Pharmacogenomics
* Understand the structure of the `CoreSet` and `PharmacoSet` classes to facilitate their use in subsequent analyses
* Download/load a `PharmacoSet` using `r PGx` or [orcestra.ca](https://orcestra.ca)
* Subset and filter a `PharmacoSet` by samples and/or treatments
* Access the molecular features, dose-response and metadata contained within the `PharmacoSet`

----

```{r global_knitr_opts, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE, include=TRUE, warning=FALSE, message=TRUE)
```

```{r load_packages, eval=TRUE, echo=FALSE, include=FALSE}
suppressPackageStartupMessages({
    library(CBWWorkshop2024)
    # library(PharmacoGx)
    library(data.table)
})
```



