---
title: "Module2: Exercises"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  Exercises corresponding to the Module 2 of the CBW workshop (see Vignettes).
---


```{r load_packages, eval=TRUE, echo=TRUE, include=FALSE}
suppressPackageStartupMessages({
  library(CBWWorkshop2024)
  library(PharmacoGx)
  library(data.table)
})
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

We will be working with the `NCI_ALMANAC_pset` object provided by this workshop.

Load the data and determine how many samples and treatments are in the dataset
```{r setup, exercise=TRUE, include=TRUE}

pset <- CBWWorkshop2024::NCI_ALMANAC_pset
pset
```

### Number of Samples

```{r n_samples, exercise=TRUE}
# after initializing the pset object, determine the number of samples and treatments
n_samples <- <YOUR CODE HERE>
n_samples
```

```{r n_samples-solution, exercise=FALSE}
n_samples <- length(sampleNames(pset))
n_samples
```


### Number of Treatments

```{r n_treatments, exercise=TRUE}
n_treatments <- <YOUR CODE HERE>
n_treatments

```


```{r n_treatments-solution, exercise=FALSE}
n_treatments <- length(treatmentNames(pset))
n_treatments
```


## Subsetting the PSet

### Metadata 

Print the first 5 rows of the sample metadata
```{r view_sample_metadata, exercise=TRUE}
# Access the sample metadata for the PSet 

# Print the first 5 rows of the sample metadata

```

```{r view_sample_metadata-solution, exercise=FALSE}
# Solution code
sample_metadata <- pset@sample
head(sample_metadata, 5)
```

### Subsetting the PSet


Subset the PSet to include only the samples treated with the drug `5-Fluorouracil`



