---
title: "Module2: Exercises"
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
  revealjs::revealjs_presentation:
    theme: solarized
    highlight: pygments
    center: false
description: >
  Exercises corresponding to the Module 2 of the CBW workshop (see Vignettes).
---


```{r global_knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, include = TRUE, warning = FALSE, message = TRUE
)
```


```{r load_packages, eval=TRUE, echo=TRUE}
suppressPackageStartupMessages({
  library(CBWWorkshop2024)
  library(PharmacoGx)
  library(data.table)
})
```


# Introduction

We will be working with the `NCI_ALMANAC_pset` object provided by this workshop.

Load the data and determine how many samples and treatments are in the dataset
```{r setup}
pset <- CBWWorkshop2024::NCI_ALMANAC_pset
pset
```

## Number of Samples

```{r n_samples-solution}

```

## Number of Treatments

```{r n_treatments-solution}


```


# Subsetting the PSet

## Metadata 

Print the first 5 rows of the sample metadata
```{r view_sample_metadata-solution}



```

## Subsetting the PSet

Subset the PSet to include only data for the sample "NCI-H322M"

```{r subset_pset-solution, warning=FALSE}

```

## Subsetting the PSet

Subset the PSet to include only data for samples where:

* The `NCI_ALMANAC.disease` is "Non-Small Cell Lung Cancer"
* and `tissueid` is NOT "Pleura"

```{r subset_pset2-solution, warning=FALSE}







```


# Subset and Aggregate

Load the `NCI_ALMANAC_TRE` object provided by this workshop.

```{r load_tre}
tre <- CBWWorkshop2024::NCI_ALMANAC_TRE
```

## Subset `TreatmentResponseExperiment`

Using the `[` operator, subset the `tre` object to include only data for sample "NCI-H322M"

```{r subset_tre2-solution}

```

## Subset `TreatmentResponseExperiment`

Using the either the `subset` method or `[` notations, 
subset the `tre` object to include only data where
* the `sampleid` starts with "NCI"
* and the `dose` is between 0.05 and 10

```{r subset_tre-solution}







```

## Aggregate `TreatmentResponseExperiment` 

* Use `aggregate` and `subset` to get all the `treatmentid`
with more than 15000 experiments

```{r tre_aggregate_exercise2}







```


## Aggregate `TreatmentResponseExperiment` and plot 

* Use `aggregate` to summarize the data by taking the mean of the `viability` values
* Aggregate over the `treatmentid`, `sampleid`, `bio_rep`, and `treatmentdose`
* Then, count the number of doses for each unique combination of `treatmentid`, `sampleid`, and `bio_rep`
* Plot the histogram of the number of doses

```{r tre_aggregate_exercise}















```


## Endoaggregation

Use `endoaggregate` to summarize the data by taking the mean of the `viability` values


## Endoaggregate `TreatmentResponseExperiment` mean viability

Use `endoaggregate` to get a `TreatmentResponseExperiment` with a new assay 
called `summary_viability` that summarizes the viability data with:
1. The `mean` 
2. The `median` 
3. The `std` 
3. The `min` 
4. The `max` 

The col names can be 
* `mean_viability`
* `median_viability` 
* etc. 

Note: you will have to determine what to aggregate over based on the data you have

```{r tre_endoaggregate_exercise}











```







